Pipeline systems and infrastructure for the efficient and open processing of Big neuroimaging Data

### 2018-09-26 paper.tex
A few comments and corrections

    \begin{abstract} Text of abstract.  
    \end{abstract} 
    \tableofcontents
    \chapter{Introduction}
        In recent years, the volume of neuroimaging data acquired has exceeded
        both the storage and computation capacity of a standard research 
        lab workstation. With the advancement of data sharing technologies, this 
        data has been made widely available, with the only factor impeding
        research on this data being access to infrastructure and software that 
        enables efficient processing on such infrastructure. Many research labs 
        do have access to high performance computing (HPC) clusters, however, 
        without efficient software to process this data, processing can take an 
        excessive amount of time. Ensuring efficient processing of data is 
        complex and likely beyond the knowledge of an average neuroimaging 
        researcher. As a result, frameworks for the efficient processing of 
        neuroimaging data need to be developed.

        Many neuroimaging workflows currently exist. Examples include
        Nipype, Pipeline System for Octave and Matlab (PSOM), LONI, SPM, FastR,
        Automated Analysis (AA) and Pydpiper\note{add citations here}. These 
        workflow engines aim to satify four criteria: 1) straightforward 
        workflow composition, 2) performance, 3) portability of the workflows 
        and 4) reproducibility of the analysis. While these workflow engines
        do tackle performance, it is mainly limited to minimizing computation
        time and does not consider data transfer times -- which are typically
        costly in Big Data settings. Therefore, these engines need to be adapted
        for the processing of neuroimaging data.

        Big Data frameworks differ from those available in neuroimaging as their
        focus is mainly on workflow composition and performance. Big Data 
        frameworks such as Apache Spark and Dask both aim to be easy-to-use,
        applicable to a wide variety of analyses and improve performance in the 
        processing of Big Data. These types of systems were primarily designed 
        to be efficient on commodity infrastructure in which transferring data
        over a network would be extremely costly.

        Currently, there exists no Big Data framework which is adapted for the
        processing of neuroimaging data. However, tools such as the Thunder 
        Project \note{add citation} allow for the processing of time-series
        neuroimaging data using Spark as the backend framework. Tools such as 
        these provide built-in functions to process the data. For users to 
        extend functionality, it is necessary for them to understand the backend
        framework. Other scientific workflow engines, such as Toil and Pegasus
        \note{add citations}, have incorporated features that enable more 
        performant workflow executions for scientific Big Data. They also enable
        the integration of Big Data frameworks as subpipelines within their 
        workflows. Despite integration with Big Data frameworks, it is not
        possible to create Big Data workflows using their API exclusively.

@@

        In section \ref{datasets}, we will briefly discuss the two types of 
        datasets that are giving rise to the neuroimaging big data. Section 
        \ref{platforms} will provide a brief introduction to the data-sharing
        and processing platforms available to researchers. We will conclude our 
        introduction on the differences between commodity and HPC infrastructure
        in section \ref{infrastructure}.

        As we wish to evaluate and compare scientific workflow frameworks with
        Big Data frameworks for the processing of neuroimaging big data, we will
        break down our analyses by the four principal features necessary for
        effective neuroimaging workflow engines. The workflow engines discussed
        will include SPM, LONI, Nipype, PSOM, FastR, AA, Pydpiper 
        for neuroimaging; Taverna, Pegasus, Make, OpenMOLE, Nextflow, Galaxy, 
        Toil for other scientific workflows; and MapReduce, Spark and Dask for
        Big Data Frameworks. Chapter \ref{workcomp} will evaluate these 
        workflows on the basis of workflow composition. In other words, we will
        look at the programming languages necessary to construct these workflows
        , how easy is it to add new tools and will these workflows be understood
        by the community. In Chapter \ref{performance}, we will proceed to 
        investigate the strategies employed by these frameworks to improve the 
        performance of their workflows. Chapter \ref{portability} will discuss
        the portability of the workflows written using these frameworks. 
        Workflow reproducibility will be then discussed in Chapter 
        \ref{reproducibility}.

@@

            As a result of the rise in datasharing in neuroscience, neuroimaging
            big data has come to fruition. The field of neuroscience has now
            reached a bottleneck where there exists sufficient data to make
            impactful studies, but the software to process these datasets have 
            not evolved with the size of the data.

            Neuroimaging big data comes in two formats: 1) large images and 2) 
            large datasets. Large images consist of a singular images ranging 
            from 100s of gigabytes to terabytes in size. A well-known example of 
            such an image is the BigBrain \note{cite here}, a histological image 
            of the brain of a healthy 69 year-old man. At its highest resolution
            , it is 1x1x20$\mu$m or 1TB in
            size. Other examples of large images can be found in electron 
            microscopy (EM), polarized light images (PLI) and micro coherence 
            tomography (microCT) \note{need to do more research here.cite}. 
            Images at 
            such high resolution are important to researchers as they provide 
            insights into aspects not otherwise detectible in images at lower 
            resolutions. However, due to their size and lack of resources 
            available to process such images, research on this data remains 
            limited.

            In contrast, large datasets are consist of many images acquired from
            multiple subjects, which are too large to be processed in their
            entirety. Notable examples of these datasets include the 
            Alzheimer's Disease Neuroimaging Initiative (ADNI), OpenfMRI, 
            the Autism Brain Imaging Data Exchange (ABIDE), the Consortium for
            Reliability and Reproducibility (CoRR), the Human Connectome Project
            (HCP) and the UK Biobank, with the later
            two being flagship examples of large datasets in neuroimaging.

@@

            To promote the advancement of knowledge of Autism Spectrum Disorders
            (ASD), ABIDE was conceived. ABIDE is an open repository of 
            preprocessed data obtained from 1112 R-fMRI datasets.
            In addition to R-fMRI, datasets also contain the corresponding 
            structural MRI and phenotypic data. As data made available in ABIDE
            is preprocessed, standardized pipelines were developed for the 
            preprocessing of the images. Popular neuroimaging processing tools
            such as CIVET, ANTs, and FreeSurfer \note{cite} were used for the 
            preprocessing of structural data, and several other pipelines were
            used for the processing of functional data.

            The CoRR was development as initiative to study and improve the 
            reliability and reproducibility of functional connectomics. The 
            data made available by the Consortium consists of test-retest data 
            obtained from 33 datasets. The repository contains data from 1629 
            subjects, in which anatomical, R-fMRI, dMRI, CBF and ASL scans are 
            available \note{obtained from website}.

            The HCP initiative aimed to characterize the human connectome of 
            1200 healthy adults, ranging from 22 to 35 years of age, by studying
            twins and their non-twin siblings. Data collected for this initiave
            includes structural MRI, R-fMRI, task fMRI and dMRI. MRI data was
            collected using a 3T scanner for all participants, with a 200 
            participants additionally undergoing scans in a higher-resolution 7T
            scanner. MEG and EEG data is available for a subset of the subjects. 
            In addition to imaging data, behavioural and genetic data were also 
            acquired. To faciliate analysis for research groups, HCP made 
            available the raw data, minimally pre-processed data and processed 
            data. Pipelines were developed by the HCP for pre- and post-
            processing. As in ABIDE, these pipelines reused many commonly-used
            neuroimaging tools such as FreeSurfer and FSL for structural data
            and FSL and other tools for functional data. The total amount of
            data made available by the HCP reaches up to nearly 1 Petabyte.

            The UK Biobank is an initiative aimed at the detection of pre-
            symptomatic markers of diseases occuring in middle-aged to elderly 
            populations. The study will gather data from questionnaires, 
            physical and cognative measures and biological samples of 500,000
            individuals. 100,000 of these individuals will additionally be 
            subjected to imaging studies. Imaging modalities used include
            structural MRI, fMRI and dMRI of the brain heart and body; low-dose
            X-ray bone and joint scans; and ultrasound of the carotid arteries.
            As with other initiatives, the UK Biobank makes available both 
            raw and processed data. This data is expected to exceed 0.2 
            Petabytes. Since processed data is provided, the UK Biobank makes 
            available it own pipelines based on tools in FSL, but also plans to
            adapt HCP pipelines from cortical surface extraction in their images
            .

            All these initiave stress the importance of data-sharing and
            large-scale studies for the advancement of science. Many researchers
            are limited to the processing of subsets of the data available as 
            they lack compute and storage resources to process the data in its
            entirety. It is also important to note that many of the pipelines 
            developed by these initiatives made use of popular neuroimaging 
            toolboxes, such as FSL, FreeSurfer, ANTS, CIVET, rather that
            creating their own. Therefore, there is a great need for bigdata
            frameworks that can leverage existing tools for the efficient 
            processing of large-scale datasets. With the availability of such 
            tools, many more research labs will be able to perform
            higher-impact studies \note{reword}.

        \section{Open-science Platforms}\label{platforms}
            Efficient processing of workflows are only but a small portion of 
            the Big Data issue in neuroscience. Sharing and storing of datasets
            and tools, secure access to and configuration of multiple computing 
            sites and provenance capture of all activities must be managed by 
            the researcher. This puts a heavy burden on researchers as they 
            require some knowledge of all these components in order to manage it
            correctly. These platforms also play a vital role in promoting 
            research through the sharing of knowledge. Users can choose to share
            their data and pipelines with others, reducing the need for others 
            to collect their own data or to develop their own workflows. It 
            also promotes robustness of results as the data and workflows will 
            be shared and tested by multiple users.

@@

            Examples of popular neuroinformatic open-science platforms include 
            CBRAIN\note{cite paper}, OpenNeuro \note{cite poster} and Neurodata
            \note{cite prepring}. OpenNeuro and NeuroData primarily focus on
            cloud deployement whereas CBRAIN is configured for the deployment
            of workflows on HPC resources. All three of these systems allow 
            users to configure, manage, share and execute their workflows with 
            the help of a web-based GUI. Integration with compute resources is
            entirely abstracted by the platforms. These platforms additionally
            provide tools to visualize their data.

@@

        \section{Infrastructure}\label{infrastructure}
            HPC infrastructure was designed durign a time when compute resources
            were scarce. As a result, resource managers for these systems employ
            strategies to efficiently use the compute resources available for 
            processing. However, with the advent of big data, there has been a 
            shift from compute-intensive to data-intensive workloads. That is, 
            with the accumulation of large datasets, data transfers have begun 
            to monopolize processing time, and with it software has been 
            designed to minimize data transfers.

            We define HPC systems as those that employ shared parallel 
            filesystem such as Lustre. This shared filesystem is the slowest of
            all storage devices as it is shared amongst all users of the system.
            Compute nodes in an HPC system may or may not contain any local 
            storage and network interconnects between nodes are typically fast.
            Resources managers used in HPC infrastructures typically employ 
            batch scheduling and execute heterogenous workflows. 
            Examples of schedulers used in HPC clusters 
            include SLURM, PBS, SGE and HTCondor \note{cite here}.

            Commodity infrastructure differs from HPC infrastructures in that
            it uses less-costly hardware. It follows a shared-nothing paradigm
            where the storage is not shared between nodes. That is all compute
            nodes are responsible for storing the data as well. File systems
            used in these clusters are distributed filesystems, such as the
            Hadoop Distributed File System (HDFS) and Alluxio \note{cite}.
            Fault-tolerance of the data is achieved through replication across
            nodes. As transferring data across the network is typically costly 
            in such infrastructures, frameworks designed to run on commodity 
            infrastructure attempt to limit the transfer of data across nodes. 
            Resource managers created specifically for these systems
            typically expect dedicated infrastructure, as is the case with 
            Yet Another Resource Negotiator (YARN), but can also execute 
            heterogeneous workflows, as can be seen with Mesos \note{cite}.

            Commodity clusters may also come in the form of cloud infrastructure
            . While the hardware employed is typically the same as other
            commodity clusters, the software used varies. Cloud infrastructures
            have the added benefit of virtualization, allowing users to 
            configure the infrastructure and software requirements to meet their
            needs. Cloud infrastructure also enables researchers to request
            dedicated resources without having to maintain these resources. One
            of the biggest limiting factors of cloud resources is the cost 
            associated with usage.

>>>>

    \begin{abstract} Text of abstract.  
    \end{abstract} 
    \tableofcontents
    \chapter{Introduction}
        In recent years, the volume of neuroimaging data acquired has exceeded
        both the storage and computation capacity of a standard research 
        lab workstation. With the advancement of data sharing technologies, this 
        data has been made widely available, \weird{with the only factor impeding
        research on this data being access to infrastructure and software that 
        enables efficient processing on such infrastructure}. Many research labs 
        do have access to high performance computing (HPC) clusters, however, 
        without efficient software to process this data, processing can take an 
        excessive amount of time. Ensuring efficient processing of data is 
        complex and likely beyond the knowledge of an average neuroimaging 
        researcher \tristan{Be careful with 'knowledge', some neuroimaging researchers are very compute-savvy}. As a result, frameworks for the efficient processing of 
        neuroimaging data need to be developed. \tristan{Add a sentence to transition from 'frameworks for efficient processing' to 'workflow engines'}

        Many neuroimaging workflow engines currently exist. Examples include
        Nipype, Pipeline System for Octave and Matlab (PSOM), LONI, SPM, FastR,
        Automated Analysis (AA) and Pydpiper\note{add citations here}. These 
        workflow engines aim to satisfy four criteria: 1) straightforward \tristan{easy? facilitated?} 
        workflow composition, 2) performance, 3) portability of the workflows 
        and 4) reproducibility of the analyses. While these workflow engines
        do tackle performance, it is mainly limited to minimizing computation
        time and does not consider data transfer times -- which are typically
        costly in Big Data settings. Therefore, these engines need to be adapted
        for the processing of neuroimaging data.

        Big Data frameworks differ from those available in neuroimaging as their
        focus is mainly on workflow composition and performance. Big Data 
        frameworks such as Apache Spark and Dask \tristan{Add references} both aim to be easy-to-use,
        applicable to a wide variety of analyses and improve performance in the 
        processing of Big Data. These types of systems were primarily designed 
        to be efficient on commodity infrastructure in which transferring data
        over a network would be extremely costly.

        Currently, there exists no Big Data framework which is adapted for the
        processing of neuroimaging data. However, tools such as the Thunder 
        Project \note{add citation} allow for the processing of time-series
        neuroimaging data using Spark as the backend framework. \weird{Tools such as 
        these} provide built-in functions to process the data. For users to 
        extend functionality, it is necessary for them to understand the backend
        framework. Other scientific workflow engines, such as Toil and Pegasus
        \note{add citations}, have incorporated features that enable more 
        performant workflow executions for scientific Big Data. They also enable
        the integration of Big Data frameworks as sub-pipelines within their 
        workflows. Despite integration with Big Data frameworks, it is not
        possible to create Big Data workflows using their API exclusively.

@@

        \tristan{Add a sentence to present the goal of the report.}
        In section~\ref{datasets}, we will briefly introduce the two 
        types of datasets that are giving rise to the neuroimaging big 
        data. Section~\ref{platforms} will provide a brief introduction 
        to the data-sharing and processing platforms available to 
        researchers. We will conclude our 
        introduction on the differences between commodity and HPC infrastructure
        in section~\ref{infrastructure}.

        As we wish to evaluate and compare scientific workflow frameworks with
        Big Data frameworks for the processing of neuroimaging big data, we will
        break down our analyses by the four principal features necessary for
        effective neuroimaging workflow engines. The workflow engines discussed
        will include SPM, LONI, Nipype, PSOM, FastR, AA, Pydpiper 
        for neuroimaging; Taverna, Pegasus, Make, OpenMOLE, Nextflow, Galaxy, 
        Toil for other scientific workflows; and MapReduce, Spark and Dask for
        Big Data Frameworks. Chapter \ref{workcomp} will evaluate these 
        workflows on the basis of workflow composition: we will % 'in other words' is on my list of expressions to avoid
        look at the programming languages necessary to construct these workflows
        , how easy is it to add new tools and will these workflows be understood
        by the community. In Chapter \ref{performance}, we will proceed to 
        investigate the strategies employed by these frameworks to improve the 
        performance of their workflows. Chapter \ref{portability} will discuss
        the portability of the workflows written using these frameworks. 
        Workflow reproducibility will be then discussed in Chapter 
        \ref{reproducibility}.

@@

            As a result of the rise in data sharing in neuroscience, neuroimaging
            Big Data has come to fruition. The field of neuroscience has now
            reached a bottleneck where there exists sufficient data to make
            impactful studies, but the software to process these datasets have 
            not evolved with the size of the data.

            Neuroimaging big data comes in two formats: 1) large images and 2) 
            large datasets. Large images consist of singular images ranging 
            from 100s of gigabytes to terabytes in size. A well-known example of 
            such an image is the BigBrain \note{cite here}, a histological image 
            of the brain of a healthy 69 year-old man. At its highest resolution
            , it is 1x1x20$\mu$m or 1TB in
            size. Other examples of large images can be found in electron 
            microscopy (EM), polarized light imaging (PLI) and micro computational
            tomography (microCT) \note{need to do more research here.cite}. 
            Images at 
            such high resolution are important to researchers as they provide 
            insights into aspects not otherwise detectible in images at lower 
            resolutions. However, due to their size and lack of resources 
            available to process such images, research on this data remains 
            limited.

            In contrast, large datasets consist of many images acquired from
            multiple subjects, which are too large to be processed in their
            entirety. Notable examples of these datasets include the 
            Alzheimer's Disease Neuroimaging Initiative (ADNI), OpenfMRI, 
            the Autism Brain Imaging Data Exchange (ABIDE), the Consortium for
            Reliability and Reproducibility (CoRR), the Human Connectome Project
            (HCP) and the UK Biobank, with the later
            two being flagship recent examples of large datasets in neuroimaging.

@@

            ABIDE was conceived to promote the advancement of knowledge 
            of Autism Spectrum Disorders (ASD). % inverted the sentence to keep symmetry with the other paragraphs
            ABIDE is an open repository of 
            preprocessed data obtained from 1112 R-fMRI datasets.
            In addition to R-fMRI, datasets also contain the corresponding 
            structural MRI and phenotypic data. As data made available in ABIDE
            is preprocessed, standardized pipelines were developed for the 
            preprocessing of the images. Popular neuroimaging processing tools
            such as CIVET, ANTs, and FreeSurfer \note{cite} were used for the 
            preprocessing of structural data, and several other pipelines were
            used for the processing of functional data.

            The CoRR \tristan{expand acronym} was developed as an initiative to study and improve the 
            reliability and reproducibility of functional connectomics. The 
            data made available by the Consortium consists of test-retest data 
            obtained from 33 datasets. The repository contains data from 1629 
            subjects, in which anatomical, R-fMRI, dMRI, CBF and ASL scans are 
            available \note{obtained from website}.

            The HCP initiative aimed to characterize the human connectome of 
            1200 healthy adults, ranging from 22 to 35 years of age, by studying
            twins and their non-twin siblings. Data collected for this initiave
            includes structural MRI, R-fMRI, task fMRI and dMRI. MRI data was
            collected using a 3T scanner for all participants, with a 200 
            participants additionally undergoing scans in a higher-resolution 7T
            scanner. MEG and EEG data is available for a subset of the subjects. 
            In addition to imaging data, behavioural and genetic data were also 
            acquired. To facilitate analysis for research groups, HCP made 
            available the raw data, minimally pre-processed data and processed 
            data. Pipelines were developed by the HCP for pre- and post-
            processing. As in ABIDE, these pipelines reused many commonly-used
            neuroimaging tools such as FreeSurfer and FSL for structural data
            and FSL and other tools for functional data. The total amount of
            data made available by the HCP reaches up to nearly 1 Petabyte.

            The UK Biobank is an initiative aimed at the detection of pre-
            symptomatic markers of diseases occurring in middle-aged to elderly 
            populations. The study will gather data from questionnaires, 
            physical and cognitive measures and biological samples of 500,000
            individuals. 100,000 of these individuals will additionally be 
            subjected to imaging studies. Imaging modalities used include
            structural MRI, fMRI and dMRI of the brain heart and body; low-dose
            X-ray bone and joint scans; and ultrasound of the carotid arteries.
            As with other initiatives, the UK Biobank makes available both 
            raw and processed data. This data is expected to exceed 0.2 
            Petabytes. Since processed data is provided, the UK Biobank makes 
            available it own pipelines based on tools in FSL, but also plans to
            adapt HCP pipelines from cortical surface extraction in their images
            .

            All these initiative stress the importance of data-sharing and
            large-scale studies for the advancement of science. Many researchers
            are limited to the processing of subsets of the data available as 
            they lack compute and storage resources to process the data in its
            entirety. It is also important to note that many of the pipelines 
            developed by these initiatives made use of popular neuroimaging 
            toolboxes, such as FSL, FreeSurfer, ANTS, CIVET, rather that
            creating their own. Therefore, there is a great need for Big Data
            frameworks that can leverage existing tools for the efficient 
            processing of large-scale datasets. With the availability of such 
            tools, many more research labs will be able to perform
            higher-impact studies \note{reword}.

        \section{Open-science Platforms}\label{platforms}
            Efficient processing of workflows are only but a small portion of 
            the Big Data issue in neuroscience. % nice transition, this section gives more context for your study
            Sharing and storing of datasets
            and tools, secure access to and configuration of multiple computing 
            sites and provenance capture of all activities must be managed by 
            the researcher. This puts a heavy burden on researchers as they 
            require some knowledge of all these components in order to manage it
            correctly. These platforms also play a vital role in promoting 
            research through the sharing of knowledge. Users can choose to share
            their data and pipelines with others, reducing the need for others 
            to collect their own data or to develop their own workflows. It 
            also promotes robustness of results as the data and workflows will 
            be shared and tested by multiple users.

@@

            Examples of popular neuroinformatics open-science platforms include 
            CBRAIN\note{cite paper}, OpenNeuro \note{cite poster} and Neurodata
            \note{cite prepring}. OpenNeuro and NeuroData primarily focus on
            cloud deployement whereas CBRAIN is configured for the deployment
            of workflows on HPC resources. All three of these systems allow 
            users to configure, manage, share and execute their workflows with 
            the help of a web-based GUI. Integration with compute resources is
            entirely abstracted by the platforms. These platforms additionally
            provide tools to visualize their data.

@@

        \section{Infrastructure}\label{infrastructure}
            HPC infrastructure was designed during a time when compute resources
            were scarce. As a result, \tristan{I disagree with 'As a result' and I don't particularly agree with the previous sentence: even if
            resources aren't scarce, they are good reasons to optimize resource managers: energy consumption, application makespan, fairness, etc. Also,
            I'm not sure if resources are less scarce now than they used to be: with the rise of computational sciences, almost every researcher now
            needs a cluster, and Compute Canada (but also other infrastructures) has been very busy.}
            resource managers for these systems employ
            strategies to efficiently use the compute resources available for 
            processing. However, with the advent of big data, there has been a 
            shift from compute-intensive to data-intensive workloads. That is, 
            with the accumulation of large datasets, data transfers have begun 
            to monopolize processing time, and with it software has been 
            designed to minimize data transfers.

            We define HPC systems as those that employ shared parallel 
            filesystem such as Lustre \tristan{and data-unaware schedulers?}.
            \tristan{You should say that this definition is 'by contrast to big data platforms' to explain where it's coming from.}
            This shared filesystem is the slowest of
            all storage devices as it is shared amongst all users of the system.
            Compute nodes in an HPC system may or may not contain any local 
            storage and network interconnects between nodes are typically fast.
            Resources managers used in HPC infrastructures typically employ 
            batch scheduling and execute heterogenous workflows. 
            Examples of schedulers used in HPC clusters 
            include SLURM, PBS, SGE and HTCondor \note{cite here}.

            Commodity infrastructure \tristan{Call it 'Big Data infrastructure?'} differs from HPC infrastructures in that
            it uses less-costly hardware. It follows a shared-nothing paradigm
            where the storage is not shared between nodes. That is, all compute
            nodes are responsible for storing the data as well. File systems
            used in these clusters are distributed filesystems, such as the
            Hadoop Distributed File System (HDFS) and Alluxio \note{cite}.
            Fault-tolerance of the data is achieved through replication across
            nodes. As transferring data across the network is typically costly 
            in such infrastructures, frameworks designed to run on commodity 
            infrastructure attempt to limit the transfer of data across nodes. 
            Resource managers created specifically for these systems
            typically expect dedicated infrastructure, as is the case with 
            Yet Another Resource Negotiator (YARN), but can also execute 
            heterogeneous workflows, as can be seen with Mesos \note{cite}.

            Commodity clusters may also come in the form of cloud infrastructure.
            While the hardware employed is typically the same as other
            commodity clusters, the software used varies. Cloud infrastructures
            have the added benefit of virtualization, allowing users to 
            configure the infrastructure and software requirements to meet their
            needs. Cloud infrastructure also enables researchers to request
            dedicated resources without having to maintain these resources. One
            of the biggest limiting factors of cloud resources is the cost 
            associated with usage.
